# 6: Analysis of Longitudinal Data
```{r}
date()
```
*This week I am learning about analysis of longitudinal data. I have done the data wrangling part and now I will do the analysis excercise.*


```{r}
#load the data
BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", header=TRUE)
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", header=TRUE)

#explore data
str(BPRS)
str(RATS)

# Factor treatment & subject
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

# Convert to long form
BPRSL <-  BPRS %>% gather(key = weeks, value = bprs, -treatment, -subject)
BPRSL <-  BPRSL %>% mutate(week = as.integer(substr(weeks,5,5)))

RATSL <- RATS %>%
  gather(key = WD, value = Weight, -ID, -Group) %>%
  mutate(Time = as.integer(substr(WD,3,4))) 

#Access the package ggplot2
library(ggplot2)

# Draw the plot
p <- ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
p

```

- In general, the rats in Group 1 has the lowest weight compared to Group 2 and 3. From this graph, we can see how rats with similar weights are grouped together, although there is a rat which is significantly heavier than others in its group in Group 2.
- All groups showed an increase in weight over time.


```{r}
# standardizing the weight variable
RATSL <- RATSL %>%
  group_by(Group) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Draw the plot
p <- ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$stdWeight), max(RATSL$stdWeight)))
p

```

- Now that the weight data has been standardized, the differences between groups are not as pronounced anymore and we can see the changes that happened in the groups more clearly.
- So far, it is not clear if there is any difference of degree of increase between groups. 
- Lighter rats in all groups seem to have the most extreme weight gain compared to heavier rats in its group.


```{r}
# Number of Time
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

- From this plot above, we can see the average increase of weight in each groups.
- Group 2 seem to have the most significant weight gain in average compared to the other groups, while Group 1 the least significant.


```{r}
# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
RATSL8S <- RATSL %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time 1-64")

```

- From the boxplot above, we see that there is an outlier in every group. We will remove those outlier.

```{r}
# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- filter(RATSL8S, (Group==1 & mean>250) | (Group==2 & mean < 590) | (Group==3 & mean>500))

# Draw a boxplot of the filtered data
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time 1-64")

```

- The outliers has been treated. By excluding the outler, the shape of the box plot changes especially in Group 2 and 3. This is because the outlier heavily affects the average of the group. 

```{r}
# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```
- We will not do a two-sample t-test as there are more than 2 groups in this data
- The baseline is highly significant, while Group is significant at 0.1 level.

```{r}
#bprs data
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

- BPRS values in the treatment group 2 seem to have more variation compared to treatment 1.
- In general, it seems that BRPS values in both group decreases in time, indicating that the treatments might be effective.

```{r}
# create a regression model RATS_reg
BPRS_reg <- lm(bprs~ week+treatment, data=BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

- The regression model shows that BPRS does significantly decrease each week, further supporting the argument that the treatments are effective.
- However, the treatment variable was not found to be significant, which means that we could not find any significant difference between the two treatments in affecting the BPRS, or in another words both treatment have similar degree of effect.


```{r}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```
- This model allows the linear regression fit for each subject to differ in intercept.

```{r}

# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)
```

-The chi-squared statistic of the likelihood ratio test between BPRS_ref and BPRS_ref1 is within the 95% level of significance, indicating a good fit for the model.

```{r}
# draw the plot 
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref1)

# Create a new column fitted to RATSL
BPRSL <- BPRSL %>%
  mutate(Fitted)

# draw the plot of RATSL
ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

- The fitted plot shows the fitted bprs values according to the model. 
- As the or